---
title: "MARBLES Vignette"
author: "Biqing Zhu"
package: "`r BiocStyle::pkg_ver('MARBLES')`"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: yes
  html_document:
    self_contained: yes
    toc: yes
vignette: >
  %\VignetteIndexEntry{MARBLES Vignette}
  %\VignettePackage{MARBLES}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: "`r file.path(system.file('data', package = 'MARBLES'), 'refs.bib')`"
---

<style type="text/css">
.smaller {
  font-size: 10px
}
</style>

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Installing MARBLES from GitHub

```{r setup, warning = FALSE, message = FALSE}
library(MARBLES)
```

# Main Function

The main function is

```{r, warning = FALSE, message = FALSE}
?run_MRF()
```

It requires three input values: (1) paraMRF: Starting value of the model parameter $\Phi$; (2) expr: A pseudo-bulk expression list for a gene. Each item is a vector of the pseudo-bulk expression across all individuals for a cell type in a condition. The list names should start with 'Cond1_' or 'Cond2_', and followed by the name of the cell type; (3) c_c: A binary cell type relationship network matrix. 1 means connected, and 0 means not connected; and (4) x_init: Initial DE status. Should be a binary vector of length n_k. 1 means DE, and 0 means EE. The usage of this function will be illustrated in the following real data application.

# Example

We obtained the mouse cortex snRNA-seq dataset from the R package `r BiocStyle::Biocpkg("muscData")` [@crowell2020muscat]. The data has already been pre-processed as well as annotated, and contains four control and four lipopolysaccharide (LPS)-treated mice.
```{r, warning = FALSE, message = FALSE}
library(muscData)
library(muscat)

mouse <- Crowell19_4vs4()
mouse <- prepSCE(mouse, 
                kid = "cluster_id", # subpopulation assignments
                gid = "group_id",  # group IDs (ctrl/stim)
                sid = "sample_id",   # sample IDs (ctrl/stim.1234)
                drop = TRUE)  # drop all other colData columns
```

Since we wanted to focus on neurons and glial cells, we only selected astrocytes, microglia, oligodendrocyte progenitor cells (OPC), oligodendrocytes, excitatory neurons, and inhibitory neurons for the downstream analyses, and only kept the first 1,000 genes for demonstration purposes. And we applied the function `aggregateData` in the R package `r BiocStyle::Biocpkg("muscat")` [@crowell2020muscat] to get cell-type-specific pseudobulk data.

```{r, warning = FALSE, message = FALSE}
pb <- aggregateData(mouse[1:1000, !mouse$cluster_id %in% c('Endothelial', "CPE cells")])
```

Then, we created a cell type relationship network based on domain knowledge to represent the similarity and cell lineage.

```{r}
## Create connection matrix (OPC connected to neurons)
celltypes <- c('Astrocytes', 'Excit. Neuron', 'Inhib. Neuron', 
               'Microglia', 'Oligodendrocytes', 'OPC')
c_c <- matrix(1, nrow = length(celltypes), ncol = length(celltypes))
rownames(c_c) <- colnames(c_c) <- celltypes
conn <- c(0, 0, 1, 0, 1, 1, 0, 0, 1, 0, 0, 1, 0, 1, 1)
c_c[lower.tri(c_c)] <- conn
c_c <- t(c_c)
c_c[lower.tri(c_c)] <- conn
c_c
```

`r BiocStyle::Biocpkg("edgeR")` [@robinson2010edger] was first applied to the pseudo-bulk data to get the DE states using `pbDS`
```{r, warning = FALSE, message = FALSE}
res_edger <- pbDS(pb, method = "edgeR", verbose = FALSE)
tbl_edger <- res_edger$table[[1]]
names(tbl_edger)
# view results for 1st cluster
k1 <- tbl_edger[[1]]
head(format(k1[, -ncol(k1)], digits = 2))
```

Next, genes with FDR corrected p value >= 0.05, and abs(logFC) <=1 were filtered out.
```{r, warning = FALSE, message = FALSE}
results_fil <- lapply(tbl_edger, function(u) {
      u <- dplyr::filter(u, p_adj.loc < 0.05 & abs(logFC) > 1)
      dplyr::arrange(u, p_adj.loc)
})
results_gene <- lapply(results_fil, function(u) u$gene)
pseudo_bulk_lst <- pb@assays@data@listData
```

In order to input the data into our model, it's structure was rearranged such that each element of the list represents a gene, and within each element, each item is a vector of the pseudo-bulk expression across all individuals for a cell type in a condition.

```{r}
pseudo_bulk_lst_new <- list()
for(gene in rownames(pseudo_bulk_lst[[1]])) {
    pseudo_bulk_lst_new[[gene]] <- list()
    for(celltype in celltypes) {
        pseudo_bulk_lst_new[[gene]][[paste0('Cond1_', celltype)]] <- 
          unlist(pseudo_bulk_lst[[celltype]][gene, 1:4])
        pseudo_bulk_lst_new[[gene]][[paste0('Cond2_', celltype)]] <- 
          unlist(pseudo_bulk_lst[[celltype]][gene, 5:8])
    }
}
str(pseudo_bulk_lst_new[[1]])
```

Finally, we ran our method with edgeR initialization.

```{r, warning = FALSE, message = FALSE}
x_mat_edger <- matrix(0, length(pseudo_bulk_lst_new), length(celltypes))
rownames(x_mat_edger) <- names(pseudo_bulk_lst_new)
colnames(x_mat_edger) <- celltypes
for (celltype in celltypes) {
   x_mat_edger[results_fil[[celltype]]$gene, celltype] <- 1
}
x_mat_edger <- lapply(seq_len(nrow(x_mat_edger)), function(i) x_mat_edger[i,])
results_edger <- mapply(run_MRF, expr = pseudo_bulk_lst_new, x_init = x_mat_edger, 
                        MoreArgs = list(paraMRF = c(0, 2), c_c = c_c), SIMPLIFY = FALSE)
```
The $\Phi$ parameter for the 10th gene can be retrieved by 

```{r}
results_edger[[10]]$phi_mat[,ncol(results_edger[[10]]$phi_mat)]
```

And the DE states is

```{r}
results_edger[[10]]$x_mat[,ncol(results_edger[[10]]$x_mat)]
```

# References